---
title: "Carbon and Tree Loss Data"
author: "Jesse Brodrick"
date: "2024-03-20"
output: html_document
---
## Time Series Regression of CO2 emssions and Tree Cover comparing the the PNW verses the US as a whole (2000-2022)

### Carbon Emission Time Series Regression in the PNW
```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)

ccd_data <- read_csv("/Users/jessebrodrick/Desktop/490Pro/data/GlblFW/CCD.csv")

# Filter for a  threshold of 50 and western states  
western_states <- c("Washington", "Oregon", "California", "Idaho", 
                    "Montana", "Nevada")

ccd_subset <- ccd_data %>%
  filter(subnational1 %in% western_states, umd_tree_cover_density_2000__threshold == 50)

### Identify the emission columns
emission_columns <- names(ccd_subset)[grepl("gfw_forest_carbon_gross_emissions_", names(ccd_subset))]

### Pivot the data to long format,
ccd_long <- ccd_subset %>%
  pivot_longer(
    cols = all_of(emission_columns),
    names_to = "Year_Emission",
    values_to = "Emissions"
  )

### Extract the year from 'Year_Emission' column
ccd_long <- ccd_long %>%
  mutate(Year = str_extract(Year_Emission, "\\d{4}")) %>%
  select(-Year_Emission) %>%
  mutate(Year = as.numeric(Year))

### Check to ensure no NAs in Year column
print(sum(is.na(ccd_long$Year)))

### Aggregate sum of by year
annual_emissions <- ccd_long %>%
  group_by(Year) %>%
  summarise(Total_Emissions = sum(Emissions, na.rm = TRUE))

### linear regression analysis
model <- lm(Total_Emissions ~ Year, data = annual_emissions)
summary(model)

# Predict using the model
annual_emissions$Predicted_Emissions <- predict(model, newdata = annual_emissions)

# Plot the actual vs predicted emissions
ggplot(annual_emissions, aes(x = Year, y = Total_Emissions)) +
  geom_line(aes(y = Predicted_Emissions), color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(y = Total_Emissions), color = "blue", size = 2) +
  labs(x = "Year", y = "Total Emissions (Mg CO2e)", title = "Actual vs Predicted Total Emissions within the PNW") +
  theme_minimal()

```

### PNW Tree Cover Loss Time Series Regression 
```{r}
# PNW Tree Cover Loss time series regression
TCL_data <- read_csv("/Users/jessebrodrick/Desktop/490Pro/data/GlblFW/CTCL1.csv")

# Filter for a  threshold of 50 and western states  
western_states <- c("Washington", "Oregon", "California", "Idaho", 
                    "Montana", "Nevada")

TCL_subset <- TCL_data %>%
  filter(subnational1 %in% western_states, threshold == 50) 

### Identify the emission columns
hloss_columns <- names(TCL_subset)[grepl("tc_loss_ha_20", names(TCL_subset))]

### Pivot the data to long format,
tcl_long <- TCL_subset %>%
  pivot_longer(
    cols = all_of(hloss_columns),
    names_to = "Year_Hectares_loss",
    values_to = "hectares_loss"
  )

### Extract the year from 'Year_hectares_loss' column
tcl_long <- tcl_long %>%
  mutate(Year = str_extract(Year_Hectares_loss, "\\d{4}")) %>%
  mutate(Year = as.numeric(Year))

### Check to ensure no NAs in Year column
print(sum(is.na(tcl_long$Year)))

### Aggregate sum of by year
annual_hloss <- tcl_long %>%
  group_by(Year) %>%
  summarise(Total_loss = sum(hectares_loss, na.rm = TRUE))

### linear regression analysis
model <- lm(Total_loss ~ Year, data = annual_hloss)
summary(model)

# Predict using the model
annual_hloss$Predicted_hectares_lossed <- predict(model, newdata = annual_hloss)

# Plot the actual vs predicted emissions
ggplot(annual_hloss, aes(x = Year, y = Total_loss)) +
  geom_line(aes(y = Predicted_hectares_lossed), color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(y = Total_loss), color = "blue", size = 2) +
  labs(x = "Year", y = "Total Tree Cover Loss (Hectares)", title = "Actual vs Predicted Total Loss within the PNW")+
  theme_minimal()
```
### Regression for Tree Cover versus Carbon Emissions (PNW)
```{r}
# Merge the two annual summaries by Year
tclco2 <- merge(annual_emissions, annual_hloss, by = "Year")

# Check the combined data
print(head(tclco2))

# Linear regression analysis with Total Emissions as the predictor of Total Loss
combined_model <- lm(Total_loss ~ Total_Emissions, data = tclco2)
summary(combined_model)

# xplore the relationship visually
ggplot(tclco2, aes(x = Total_Emissions, y = Total_loss)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(x = "Total Emissions (Mg CO2e)", y = "Total Tree Cover Loss (Hectares)",
       title = "Relationship between Carbon Emissions and Tree Cover Loss within the PNW") +
  theme_minimal()
```
## Compared Regression of Tree Cover versus Carbon Emissions for PNW and the Entire US
### US Tree Cover Time Series Regression
```{r}
# Tree Cover for Entire US
TCL_USsubset <- TCL_data %>%
  filter(threshold == 50) 

### Identify the emission columns
UShloss_columns <- names(TCL_USsubset)[grepl("tc_loss_ha_20", names(TCL_USsubset))]

### Pivot the data to long format,
tcl_USlong <- TCL_USsubset %>%
  pivot_longer(
    cols = all_of(UShloss_columns),
    names_to = "Year_USHectares_loss",
    values_to = "UShectares_loss")

### Extract the year from 'Year_UShectares_loss' column
tcl_USlong <- tcl_USlong %>%
  mutate(Year = str_extract(Year_USHectares_loss, "\\d{4}")) %>%
  mutate(Year = as.numeric(Year))

### Check to ensure no NAs in Year column
print(sum(is.na(tcl_USlong$Year)))

### Aggregate sum of by year
annual_USloss <- tcl_USlong %>%
  group_by(Year) %>%
  summarise(Total_loss = sum(UShectares_loss, na.rm = TRUE))

### linear regression analysis
USmodel <- lm(Total_loss ~ Year, data = annual_USloss)
summary(USmodel)

# Predict using the model
annual_USloss$Predicted_hectares_lossed <- predict(model, newdata = annual_USloss)



```
### US Carbon Emission Time Series Regression  
```{r}
ccd_USsubset <- ccd_data %>%
  filter(umd_tree_cover_density_2000__threshold == 50)

### Identify the emission columns
USemission_columns <- names(ccd_USsubset)[grepl("gfw_forest_carbon_gross_emissions_", names(ccd_USsubset))]

### Pivot the data to long format,
USccd_long <- ccd_USsubset %>%
  pivot_longer(
    cols = all_of(USemission_columns),
    names_to = "US_Year_Emission",
    values_to = "US_Emissions"
  )

### Extract the year from 'US_Year_Emission' column
USccd_long <- USccd_long %>%
  mutate(Year = str_extract(US_Year_Emission, "\\d{4}")) %>%
  select(-US_Year_Emission) %>%
  mutate(Year = as.numeric(Year))

### Check to ensure no NAs in Year column
print(sum(is.na(USccd_long$Year)))

### Aggregate sum of by year
USannual_emissions <- USccd_long %>%
  group_by(Year) %>%
  summarise(USTotal_Emissions = sum(US_Emissions, na.rm = TRUE))

### linear regression analysis
model <- lm(USTotal_Emissions ~ Year, data = USannual_emissions)
summary(model)

# Predict using the model
USannual_emissions$USPredicted_Emissions <- predict(model, newdata = USannual_emissions)

# Plot the actual vs predicted emissions
ggplot(USannual_emissions, aes(x = Year, y = USTotal_Emissions)) +
  geom_line(aes(y = USPredicted_Emissions), color = "red", linetype = "dashed", linewidth = 1) +
  geom_point(aes(y = USTotal_Emissions), color = "blue", size = 2) +
  labs(x = "Year", y = "Total Emissions (Mg CO2e)", title = "Actual vs Predicted Total Emissions within the PNW") +
  theme_minimal()
```

## Tree Cover Loss vs. CO2 Emissions: PNW vs. US Plot
```{r}
# Merge the PNW and US tree cover loss and emissions data

# PNW data
pnw_merged <- merge(annual_hloss, annual_emissions, by = "Year")
pnw_merged$Region <- "PNW"

# US data
us_merged <- merge(annual_USloss, USannual_emissions, by = "Year")
us_merged$Region <- "US"

summary(lm(Total_loss ~ USTotal_Emissions, data = us_merged))
        
# Making names consistent
colnames(pnw_merged)[colnames(pnw_merged) == "Total_loss"] <- "Total_Loss"
colnames(pnw_merged)[colnames(pnw_merged) == "Total_Emissions"] <- "Total_Emissions"

colnames(us_merged)[colnames(us_merged) == "Total_loss"] <- "Total_Loss"
colnames(us_merged)[colnames(us_merged) == "USTotal_Emissions"] <- "Total_Emissions"
colnames(us_merged)[colnames(us_merged) == "USPredicted_Emissions"] <- "Predicted_Emissions"

#  combining again
combined_data <- rbind(pnw_merged, us_merged)

# Check joined data
head(combined_data)

# Plot the combined data
ggplot(combined_data, aes(x = Total_Emissions, y = Total_Loss, color = Region)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Total CO2 Emissions (Mg CO2e)", y = "Total Tree Cover Loss (Hectares)",
       title = "Tree Cover Loss vs. CO2 Emissions: PNW vs. US") +
  scale_color_manual(values = c("PNW" = "green", "US" = "blue")) +
  theme_minimal()
```

## Discussion: 

#### When looking at the scatterplot comparing total tree cover loss in hectares to total CO2 emissions for the PNW and US. Each point represents an aggregated data point over time. In both cases, there appears to be a positive correlation between CO2 emissions and tree cover loss, indicating that higher emissions are associated with greater loss of tree cover. 

#### Specifically for the PNW, the slope of the trend line is less steep, suggesting a smaller increase in tree cover loss for each unit of CO2 emissions when compared to the US overall. However, in terms of scale the data points and trend are closely located to each other. This means that the 5 states within the PNW chosen represent a large portion within the US as a whole. Additionally, the PNW data values appear to be more tightly clustered around the trend line, indicating a more consistent relationship between emissions and tree cover loss within this region. For the US, the trend line is steeper, which could indicate a larger increase in tree cover loss for each unit of CO2 emissions. The data points are more spread out, which most likely reflects the broader range of conditions influencing this relationship across the entire country. The difference in their correlation can be seen with the R^2 value of the PNW being 0.86 compared to the lower US R^2 value of 0.2. These trends could be influenced by a variety of factors, including regional differences in land use, forest management practices, and industrial activities. The graph suggests that region-specific factors significantly influence the relationship between CO2 emissions and tree cover loss.

####  The relationship between CO2 emissions and time for the United States at an aggregated census level over a 20-year period indicate that there is a statistically significant upward trend in CO2 emissions over the 20 years covered by the dataset, with a substantial portion of the variation in emissions being explained by the passage of time. However, given the size of the residual error, other factors not included in the model likely also influence CO2 emissions.

#### The relationship between total tree cover loss and time at an aggregated census level over a 20-year period indicate that there is not enough strong evidence for a time trend in tree cover loss, and other variables not included in the model may better explain the changes in tree cover loss. 









