---
title: "Local Moran's I Clustering Analysis"
author: "Jesse Brodrick"
date: "2024-03-14"
output: html_document
html_document:
    css: html-md-01.css
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: false
    collapsed: no
---

## 20 year Tree Cover Gain Anlaysis with Spatial Distribution and Clustering


### Reading in libraries and dataset
```{r}
library(tidyverse)
library(sf)
library(ggplot2)
library(RColorBrewer)
library(tigris)
library(ggplot2)
library(classInt)
library(spdep)
library(tmap)
getwd()

df <- read_csv("/Users/jessebrodrick/Desktop/490Pro/data/GlblFW/CTCL1.csv")

```

## Basemap and projection
```{r}
# Options to return tigris objects as sf objects
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

# Obtain state boundaries
states_sf <- states(cb = TRUE)

# Obtain county boundaries
counties_sf <- counties(cb = TRUE)

# Filter for western states and their counties
western_states <- c("Washington", "Oregon", "California", "Idaho", 
                    "Montana", "Nevada")

# Filter state and counties for western states sf
wus_states_sf <- states_sf[states_sf$NAME %in% western_states, ]

wus_counties_sf <- counties_sf[counties_sf$STATE_NAME %in% western_states, ]

 

# Projection
laea = st_crs("+proj=laea +lat_0=30 +lon_0=-100") # Lambert Azimuthal Equal Area

wus_states_sf_pro = st_transform(wus_states_sf, laea)

wus_counties_sf_pro = st_transform(wus_counties_sf, laea)
```


## Subset and merge Global Forest Watch data and projected US Census boundaries

### Global Forest watch provides thresholds or amount of tree canopy coverage (density of 30m resolution) for 0, 10, 20, 30, 50, and 75 percent for 20 years of forest cover gain/loss in hecatares. 

```{r}
# Subset data based on the western states with a canopy density threshold of 50
df_subset <- df[df$subnational1 %in% western_states, ] %>% 
                  filter(threshold == 50)

head(df_subset)

# Merging projected SF to GFW data frame
wus_gbl_merge <- wus_counties_sf_pro %>%
  left_join(df_subset, by = c("STATE_NAME" = "subnational1", "NAME" = "subnational2"))

# Replacing NA's with zero
wus_gbl_merge <- wus_gbl_merge %>%
  mutate(gain_2000_2020_ha = replace_na(gain_2000_2020_ha, 0))


```


## Local Moran's I for Tree Cover Gain

```{r}

nb <- poly2nb(wus_gbl_merge, queen=TRUE)

nb[1]

lw <- nb2listw(nb, style="W", zero.policy=TRUE)

lw$weights[1]

gain.lag <- lag.listw(lw, wus_gbl_merge$gain_2000_2020_ha)
gain.lag

plot(gain.lag ~ wus_gbl_merge$gain_2000_2020_ha, pch=16, asp=1)
M1 <- lm(gain.lag ~ wus_gbl_merge$gain_2000_2020_ha)
abline(M1, col="blue")

coef(M1)[2]

I <- moran(wus_gbl_merge$gain_2000_2020_ha, lw, length(nb), Szero(lw))[1]
I

moran.test(wus_gbl_merge$gain_2000_2020_ha,lw, alternative="greater")

####There is less than a 6% likelihood that the observed pattern could be the result of random chance

#p-value of 2.2e-16 suggesting that there would be a an extremely low chance of being wrong in rejecting the null hypothesis or that there is a 0.000.. chance that our observed pattern is consistent with a random process 

MC<- moran.mc(wus_gbl_merge$gain_2000_2020_ha, lw, nsim=999, alternative="greater")

# View results (including p-value)
MC

# Plot the Null distribution (note that this is a density plot instead of a histogram)
plot(MC)

#The curve shows the distribution of Moran I values we could expect had the incomes been randomly distributed across the counties. Note that our observed statistic, 0.54995, falls  to the right of the distribution suggesting that the income values are clustered (a positive Moran’s I value suggests clustering whereas a negative Moran’s I value suggests dispersion).

# A visual used  assess how “typical” or “atypical” your pattern may be relative to a randomly distributed pattern is to plot your observed pattern alongside a few simulated patterns generated under the null hypothesis
set.seed(131)
wus_gbl_merge$rand1 <- sample(wus_gbl_merge$gain_2000_2020_ha, length(wus_gbl_merge$gain_2000_2020_ha), replace = FALSE)
wus_gbl_merge$rand2 <- sample(wus_gbl_merge$gain_2000_2020_ha, length(wus_gbl_merge$gain_2000_2020_ha), replace = FALSE)
wus_gbl_merge$rand3 <- sample(wus_gbl_merge$gain_2000_2020_ha, length(wus_gbl_merge$gain_2000_2020_ha), replace = FALSE)

tm_shape(wus_gbl_merge) + tm_fill(col=c("gain_2000_2020_ha", "rand1", "rand2", "rand3"),
                      style="quantile", n=8, palette="Greens", legend.show = FALSE) +
              tm_facets( nrow=1)
# Calculate Local Moran's I
local_morans_i <- localmoran(wus_gbl_merge$gain_2000_2020_ha, lw)

# Convert to a dataframe for easier manipulation
local_morans_df <- as.data.frame(local_morans_i)

# Add the Local Moran's I values and p-values to your spatial data frame
wus_gbl_merge$localI <- local_morans_df[,1]
wus_gbl_merge$p_value <- local_morans_df[,5]

# Classify into hot spots, cold spots, etc., based on Local Moran's I and p-values
# Here we consider hot spots as positive local I values with significant p-values (e.g., p < 0.05)
# and cold spots as negative local I values with significant p-values.
wus_gbl_merge$spot_type <- case_when(
  wus_gbl_merge$localI > 0 & wus_gbl_merge$p_value < 0.05 ~ "Hot Spot",
  wus_gbl_merge$localI < 0 & wus_gbl_merge$p_value < 0.05 ~ "Cold Spot",
  TRUE ~ "Not Significant"
)

# Visualize the hot and cold spots
tm_shape(wus_gbl_merge) +
  tm_fill(col = "spot_type", palette = c("Hot Spot" = "red", "Cold Spot" = "blue", "Not Significant" = "grey"), 
          title = "Spot Type") +
  tm_borders() +
  tm_layout(legend.title.size = 0.7, legend.text.size = 0.6) +
  tm_legend(position = c("right", "bottom"))

```
## Local Moran's I for Tree Cover Loss
```{r}
# Merging projected SF to GFW data frame
wus_loss_merge <- wus_counties_sf_pro %>%
  left_join(df_subset, by = c("STATE_NAME" = "subnational1", "NAME" = "subnational2"))
  
# Replacing NA's with zero
total_loss_by_county <- wus_loss_merge %>%
  mutate(across(starts_with("tc_loss_ha_"), as.numeric, .names = "numeric_{.col}")) %>%
  rowwise() %>%
  mutate(loss_2000_2020_ha = sum(c_across(starts_with("numeric_tc_loss_ha_")), na.rm = TRUE)) %>%
  ungroup() %>%
  select(-starts_with("numeric_tc_loss_ha_")) # Removing temporary numeric columns



nb1 <- poly2nb(total_loss_by_county, queen=TRUE)

nb1[1]

lw1 <- nb2listw(nb1, style="W", zero.policy=TRUE)

lw1$weights[1]

loss.lag <- lag.listw(lw, total_loss_by_county$loss_2000_2020_ha)
loss.lag

plot(loss.lag ~ total_loss_by_county$loss_2000_2020_ha, pch=16, asp=1)
M1 <- lm(loss.lag ~ total_loss_by_county$loss_2000_2020_ha)
abline(M1, col="blue")

coef(M1)[2]

I <- moran(total_loss_by_county$loss_2000_2020_ha, lw1, length(nb), Szero(lw1))[1]
I

moran.test(total_loss_by_county$loss_2000_2020_ha,lw1, alternative="greater")

####There is less than a 6% likelihood that the observed pattern could be the result of random chance

#p-value of 2.2e-16 suggesting that there would be a an extremely low chance of being wrong in rejecting the null hypothesis or that there is a 0.000.. chance that our observed pattern is consistent with a random process 

MC<- moran.mc(total_loss_by_county$loss_2000_2020_ha, lw1, nsim=999, alternative="greater")

# View results (including p-value)
MC

# Plot the Null distribution (note that this is a density plot instead of a histogram)
plot(MC)

#The curve shows the distribution of Moran I values we could expect had the incomes been randomly distributed across the counties. Note that our observed statistic, 0.54995, falls  to the right of the distribution suggesting that the income values are clustered (a positive Moran’s I value suggests clustering whereas a negative Moran’s I value suggests dispersion).

set.seed(131)
total_loss_by_county$rand1 <- sample(total_loss_by_county$loss_2000_2020_ha, length(total_loss_by_county$loss_2000_2020_ha), replace = FALSE)
total_loss_by_county$rand2 <- sample(total_loss_by_county$loss_2000_2020_ha, length(total_loss_by_county$loss_2000_2020_ha), replace = FALSE)
total_loss_by_county$rand3 <- sample(total_loss_by_county$loss_2000_2020_ha, length(total_loss_by_county$loss_2000_2020_ha), replace = FALSE)

tm_shape(total_loss_by_county) + tm_fill(col=c("loss_2000_2020_ha", "rand1", "rand2", "rand3"),
                      style="quantile", n=8, palette="Reds", legend.show = FALSE) +
              tm_facets( nrow=1)

# Calculate Local Moran's I for tree cover loss
local_morans_i_loss <- localmoran(total_loss_by_county$loss_2000_2020_ha, lw1)

# Incorporate Local Moran's I stats into the total_loss_by_county dataframe
total_loss_by_county$localI <- local_morans_i_loss[,1]  # Local Moran's I value
total_loss_by_county$p.value <- local_morans_i_loss[,5]  # P-value

# Define hot (High-High) and cold (Low-Low) spots based on Local Moran's I and p-value
total_loss_by_county$hot_cold_zone <- ifelse(total_loss_by_county$localI > 0 & total_loss_by_county$p.value < 0.05, "Hot Spot",
                                             ifelse(total_loss_by_county$localI < 0 & total_loss_by_county$p.value < 0.05, "Cold Spot", "Not Significant"))

# Visualize Hot and Cold Zones with tmap
tm_shape(total_loss_by_county) +
  tm_fill(col = "hot_cold_zone",
          palette = c("Hot Spot" = "red", "Cold Spot" = "blue", "Not Significant" = "grey"),
          title = "Tree Cover Loss Hot and Cold Zones") +
  tm_borders() +
  tm_legend(position = c("right", "bottom")) +
  tm_layout(frame = FALSE)
```


