---
title: "Tree Cover change in the Western U.S (50% threshold in 30m resolution)"
author: "Jesse Brodrick"
date: "2024-03-20"
output: html_document
---
## 20 year Tree Cover Loss Mapping Anlaysis with Global Forest Watch Data 

### Loading in libraries
```{r}
library(tidyverse)
library(sf)
library(ggplot2)
library(RColorBrewer)
library(tigris)
library(ggplot2)
library(classInt)
library(spdep)
library(tmap)
getwd()

ctcldf <- read_csv("/Users/jessebrodrick/Desktop/490Pro/data/GlblFW/CTCL1.csv")
```

### Simple features and projection
```{r}
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)

# Obtain state boundaries
states_sf <- states(cb = TRUE)

# Obtain county boundaries
counties_sf <- counties(cb = TRUE)

# Filter for western states and their counties
western_states <- c("Washington", "Oregon", "California", "Idaho", 
                    "Montana", "Nevada")
# Filter state and counties for western states sf
wus_states_sf <- states_sf[states_sf$NAME %in% western_states, ]

wus_counties_sf <- counties_sf[counties_sf$STATE_NAME %in% western_states, ]

 

# Projection
laea = st_crs("+proj=laea +lat_0=30 +lon_0=-100") # Lambert Azimuthal Equal Area

wus_states_sf_pro = st_transform(wus_states_sf, laea)

wus_counties_sf_pro = st_transform(wus_counties_sf, laea)
```


### Subsetting Data
```{r}
# Subset data based on the western states with a canopy density threshold of 50
Wustclsubset <- ctcldf[ctcldf$subnational1 %in% western_states, ] %>% 
                  filter(threshold == 50)

# Combine state and county names into a new column for graph visualization 
Wustclsubset1 <- Wustclsubset %>%
  mutate(region = paste(subnational1, subnational2, sep = ", "))


# Calculate total tree cover loss for each county
total_loss_by_county1 <- Wustclsubset1 %>%
  group_by(region) %>%
  summarize(Total_Loss1 = sum(tc_loss_ha_2001:tc_loss_ha_2022, na.rm = TRUE)) %>%
  arrange(desc(Total_Loss1))

# Top 10 counties with the highest total tree cover loss
top_counties_loss1 <- total_loss_by_county1 %>%
  top_n(10, Total_Loss1)



# Plotting
ggplot(top_counties_loss1, aes(x = reorder(region, Total_Loss1), y = Total_Loss1)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Top 10 Counties by Total Tree Cover Loss (2001-2022)",
       x = "County",
       y = "Total Tree Cover Loss (ha)")


```

### Assess spatial distribution of Canopy Cover Gain and Descriptive statistics

#### Plotting top tree loss counties

```{r}
# Summary statistics

# Merging projected SF to GFW data frame
wus_loss_merge <- wus_counties_sf_pro %>%
  left_join(Wustclsubset, by = c("STATE_NAME" = "subnational1", "NAME" = "subnational2"))
  
# Replacing NA's with zero
total_loss_by_county <- wus_loss_merge %>%
  mutate(across(starts_with("tc_loss_ha_"), as.numeric, .names = "numeric_{.col}")) %>%
  rowwise() %>%
  mutate(loss_2000_2020_ha = sum(c_across(starts_with("numeric_tc_loss_ha_")), na.rm = TRUE)) %>%
  ungroup() %>%
  select(-starts_with("numeric_tc_loss_ha_")) # Removing temporary numeric columns

head(wus_loss_merge)

head(total_loss_by_county)




summary(total_loss_by_county$loss_2000_2020_ha)

# Histogram 
ggplot(total_loss_by_county, aes(x = loss_2000_2020_ha)) +
  geom_histogram(binwidth = 200, fill = "skyblue", color = "black") +
  labs(title = "Histogram", x = "Tree Cover Gain (ha)", y = "Frequency")

# Boxplot highlighting PNW outliers by state
PNW <- c("Washington", "Oregon", "California", "Idaho", "Montana")

PNW_SF <- total_loss_by_county[total_loss_by_county$STATE_NAME %in% PNW, ]

head(PNW_SF)

quantile(total_loss_by_county$loss_2000_2020_ha)

ggplot(PNW_SF, aes(x = STATE_NAME, y = loss_2000_2020_ha)) +  
  geom_boxplot(width = 0.5, outlier.shape = 8, outlier.color = "red", outlier.size = 3) + 
  labs(title = "Boxplot of Tree Cover Gain by State", y = "Tree Cover Gain (ha)", x = "") +
  facet_wrap( ~ STATE_NAME, scales = "free_x", ncol = 2) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "none",
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank()) +
  coord_cartesian(ylim = c(-0, 100000))
```

#### Visualizing on a plot
```{r}

# Merging projected SF to GFW data frame
wus_loss_merge <- wus_counties_sf_pro %>%
  left_join(Wustclsubset, by = c("STATE_NAME" = "subnational1", "NAME" = "subnational2"))
  
# Replacing NA's with zero
total_loss_by_county <- wus_loss_merge %>%
  mutate(across(starts_with("tc_loss_ha_"), as.numeric, .names = "numeric_{.col}")) %>%
  rowwise() %>%
  mutate(loss_2000_2020_ha = sum(c_across(starts_with("numeric_tc_loss_ha_")), na.rm = TRUE)) %>%
  ungroup() %>%
  select(-starts_with("numeric_tc_loss_ha_")) # Removing temporary numeric columns

head(wus_loss_merge)

head(total_loss_by_county)

# Classifying data for visualization using natural jenks method
ctpnts <- classIntervals(total_loss_by_county$loss_2000_2020_ha, n = 6, style = "jenks")

cols <- brewer.pal(6, "YlOrRd")

colors <- colorRampPalette(cols)(length(ctpnts$brks)-1)

# Plotting 
ggplot(total_loss_by_county) +
  geom_sf(aes(fill = loss_2000_2020_ha)) +  
  scale_fill_gradientn(colors = colors,
                      breaks = ctpnts$brks[-length(ctpnts$brks)],
                       labels = rev(paste(round(ctpnts$brks[-length(ctpnts$brks)], 1), "to",
                                      round(ctpnts$brks[-1], 1))),
                       limits = range(total_loss_by_county$loss_2000_2020_ha)) +
  labs(title = "20 Year Forest Cover Loss (50% threshold)", fill = "Loss (hectares)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "right",
        legend.background = element_rect(fill = "white", colour = "black"),
        legend.key = element_blank(), 
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14)) +
  guides(fill = guide_legend(override.aes = list(fill = rev(colors)))) +
   geom_sf(data = wus_states_sf_pro, fill = NA, color = "black", size = 1.5) 



```

### Variation of canopy cover Loss

#### Creating standard distribution table to visualize amount of object values within sd break points

```{r}
# descriptive stats
mean_loss <- mean(total_loss_by_county$loss_2000_2020_ha, na.rm = TRUE)
str(total_loss_by_county$loss_2000_2020_ha)
sd_loss <- sd(total_loss_by_county$loss_2000_2020_ha, na.rm = TRUE)

labels <- c("<-2 SD", "-2 SD to -1 SD", "-1 SD to mean", "mean to 1 SD", " 1 SD to 2 SD", ">2 SD")

breakpoints <- c(-Inf, mean_loss - 2*sd_loss, mean_loss - sd_loss, mean_loss, mean_loss + sd_loss, mean_loss + 2*sd_loss, Inf)

total_loss_by_county$categories <- cut(total_loss_by_county$loss_2000_2020_ha, breaks = breakpoints, labels = labels)

table(total_loss_by_county$categories)
```


#### Plotting distribution of canopy cover change map

```{r}
breakpoints <- c(-Inf, mean_loss, mean_loss + sd_loss, mean_loss + 2*sd_loss, Inf)
labels <- c("Below Average (<-2 SD to mean)", "Average (mean to 1 SD)", "Above Average(1 SD to 2 SD)", "Exceptionally High(>2 SD)")


colors <- brewer.pal(4, "YlOrRd")

# Create a factor with labels based on cut points
total_loss_by_county$loss_category <- cut(total_loss_by_county$loss_2000_2020_ha, 
                                   breaks = breakpoints, 
                                   labels = labels)

# Plot
ggplot(total_loss_by_county) +
  geom_sf(aes(fill = loss_category)) +  
  scale_fill_manual(values = colors) +
  labs(title = "Tree Cover Gain Distribution (50% threshold)",
       fill = "Tree Cover Gain (SD)") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "right",
        legend.background = element_rect(fill = "white", colour = "black"),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 10, face = "bold")) +
   geom_sf(data = wus_states_sf_pro, fill = NA, color = "black", size = 1.5) 


```

